The Power of Unbiased Recursive Partitioning:\\ A Unifying View of CTree, MOB, and GUIDE

### 2020-01-26 pkg/partykit/inst/guideliketest/draft/draft.Rnw
added references
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2963 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 3382c44655d0a3c6388be51f8c9bbc8eda620998

Linear models as presented in this study have been chosen as they are highly relevant in practice,
allow for simple illustrations, and theoretical insights are available for the testing strategies.
However, the results can be easily extended to a wide variety of other models where the introduced building
blocks can be applied in the same way. Specifically, it has been shown theoretically that certain
changes in the parameters do not lead to shifts in the residuals (but in other components of the
scores). \cite{Ploberger+Kraemer:1992} showed that residual-based tests can detect a change in the parameters
of a linear model only if it also causes a shift in the expected value $\mathbb{E}(Y)$. 
This is not the case if changes are orthogonal to the mean regressor which in our case is $(1,0)^{\top}$.
Consequently, if only the slope $\beta_1$ but not the intercept $\beta_0$ changes, the shift
is of type $(0, \delta)^\top$ and thus orthogonal to the mean regressor. Due to this residual-based
tests as in GUIDE break down and do not have power to detect this. Note that this situation does not
have to be rare in practice: Especially for binary regressor variables (e.g., as in treatment-subgroup
investigations) it can easily occur \citep[see Figure~2 in][for an illustration]{Loh+He+Man:2015}.

>>>>

Linear models as presented in this study have been chosen as they are highly relevant in practice,
allow for simple illustrations, and theoretical insights are available for the testing strategies.
However, the results can be easily extended to a wide variety of other models where the introduced building
blocks can be applied in the same way. Specifically, it has been shown theoretically that certain
changes in the parameters do not lead to shifts in the residuals (but in other components of the
scores). \cite{Ploberger+Kraemer:1992} showed that residual-based tests can detect a change in the parameters
of a linear model only if it also causes a shift in the expected value $\mathbb{E}(Y)$. 
This is not the case if changes are orthogonal to the mean regressor which in our case is $(1,0)^{\top}$.
Consequently, if only the slope $\beta_1$ but not the intercept $\beta_0$ changes, the shift
is of type $(0, \delta)^\top$ and thus orthogonal to the mean regressor. Due to this residual-based
tests as in GUIDE break down and do not have power to detect this. Note that this situation does not
have to be rare in practice: Especially for binary regressor variables (e.g., as in treatment-subgroup
interactions, see \citealp{Doove+Dusseldorp+VanDeun:2014}) it can easily occur \citep[see Figure~2 in][for an illustration]{Loh+He+Man:2015}.
%% FIXME: add reference to Seibold+Hothorn+Zeileis:2019
Testing strategies being capable of detecting all kinds of changes including those orthogonal to the 
mean are also employed in an extension of (generalised) linear model trees introduced by 
\cite{Seibold+Hothorn+Zeileis:2019}
which allows for selected model parameters to be kept fixed at a globally estimated value while only 
changes in the remaining parameters are considered for splitting.
